Linking microbial communities to ecosystem functions: what we can learn from genotype-phenotype mapping in organisms

### 2019-12-10 manuscript/response_rev1.Rmd

---
output: pdf_document
---

Response to Reviewers

Dear Dr. Martiny and Reviewers,

Thank you so much for your thoughtful and helpful reviews.

We have revised the manuscript to reflect your comments and have provided
specific responses to your comments below.

>Referee: 1
>
>Comments to Author(s) The manuscript ``Linking microbial communities to
>ecosystem functions: what we can learn from genotype-phenotype mapping in
>organisms'' by Morris et al proposes that integrating traditional
>biodiversity-ecosystem function research with genotype-phenotype mapping
>approach and suggests this  could provide the new perspective microbial BEF
>studies.
>
>I believe paper has a value in terms of expanding our tools to validate data
>from microbial BEF -particularly, identifying potential key taxa.  I have two
>major concerns;
>
>1. The paper set out with the projection that current approaches of testing
>   microbial BEF are not good enough. I do not see evidence how this approach
>   is better than on any existing approach (which link microbial diversity,
>   community composition with ecosystem functions). The manuscript does provide
>   an additional tool to potential identity of key taxa which are involved in
>   specific functions. Introduction should reflect that.

The intention of this paper is not to show that previous approaches are bad or
that this approach is “better.” Instead, the goal of this paper is to lay out
how this question has been addressed in the past, present a new way of framing
the question, and then demonstrate one example of how that new framing might be
used. We have revised throughout the paper to try to reflect that goal more
accurately. In particular, we have toned down the critique of past BEF work and
made the introduction clearer about the fact that we have learned a lot from
BEF research and that this is a new way of thinking about the problem.

>2. Paper has not done justice to recent progress (since 2015), we made in
>   demonstrating microbial BEF from field to global scales. Few examples
>   provided below 
>
>   Jing et al (2015). Nature Communications 
>
>   Delgado-Baquerizo et al., (2016) Nature Communication 
>
>   Delgado-Baquerizo et al., (2017) New Phytologist 
>
>   There are several more!

We have incorporated these references into the brief review of BEF research and
our review spans the literature from 2000 to 2018.

>Other comments
>1. L26-36. There have been number of recent papers published on this topic
>   which provide evidence of microbial BEF. Please see above.  
We have edited the text in this section to include these papers.
>2. L79-81. But there are other papers which provide support for microbial BEF. 
>3. L91-95. Here and several times later on- authors used one paper to suggest
>   that either there is no or extremely limited evidence of microbial BEF.
>   However, authors should  balance that there are multiple papers which
>   provided strong evidence in support of microbial BEF in field conditions
Here and elsewhere we have made it clear what aspect of the BEF literature
that we are critiquing and we have separated the critique from the review of
literature in support of microbial BEF.
4. L290-292. Can phage treatment be useful here?  Phage treatment is a possible
   approach and we have added it with a citation.
5. Reference list needs be carefully formatted to meet the criteria of the
   journalisms.

We have added the papers above and emphasized the support for microbial BEF,
particularly in the first paragraph of the Intro and in section 1.1. We have
also softened the language surrounding what we don’t know to more accurately
capture the state of the literature. We have also been more specific about what
questions remain open, particularly in section 1.2. We have updated the
citation format to match the Phil Trans B style.

Referee: 2

Comments to Author(s) Overview: For decades, ecologists have been developing
theoretical and empirical approaches for understanding the contribution of
species to ecosystem function. While advances have been made, relationships can
still be hard to detect, especially in microbial communities where there are
many taxa that are interacting and contributing in complex ways to different
functions. The premise of the Morris et al. manuscript is that ideas and
approaches borrowed from population genomics can be used to better understand
contributions of taxa to ecosystem function. The authors argue that that
biodiversity ecosystem functioning (BEF) relationships in microbial systems
have shown promise, but that they are often clouded by covariation between
environmental filters and ecosystem function as well as correction levels that
do not match the data. Specifically, methods from genotype-phenotype mapping
provide three insights, “agnostic” mapping, correction for multiple comparison,
and the ability to account for covariation between variables such as
environmental factors, geography, and community composition. The authors use an
example of methane oxidation in soil cores where previous analysis showed no
link between microbial community structure and function. After correcting for
multiple comparisons and accounting for covariation, their approach identified
six uniquely associated taxa. The authors conclude that genotype-phenotype
mapping can improve our ability to link microbial taxa with ecosystem functions
independent of environmental filters and other factors that covary with
community structure.

Major concerns:

-- The premise of using genotype-phenotype mapping to understand BEF is
intriguing. So, I was excited to read this manuscript. A major concern is that
I thought that authors spent too much space in the relatively short manuscript
focusing on the shortcomings of BEF research (sections in Abstracts and 1.1 –
1.3) relative to what I would have liked to have learned about
genotype-phenotype mapping (1.4). The latter needs to be better developed in
order for the manuscript to have its desired effect, which I assume is to help
researchers working on BEF to incorporate genotype-phenotype mapping approaches
to their research. I’m afraid that in the current version, the manuscript falls
short of providing the necessary background and justification.

We have shortened the review of the BEF literature and have tried to be more
clear and concise about what aspect of BEF we are critiquing, insofar as it
helps develop the mapping framework. We have also tried to unpack some of the
analogies that we use and provide examples of what they might like like in
microbial communities. See, for example, the new text in sections 2.1 and 2.2.

-- I sort of like the section advocating for an “agnostic” approach given that
we often don’t know a priori which taxa are contributing directly or indirectly
to a function. However, I was more skeptical about attributing multiple
comparisons to a genotype-phenotype approach. For example, it is well known
that Bonferroni corrections are severe and alternatives have been in place for
decades.  For example, without correction for false discovery rate (FDR), the
author report that 460 taxa contributed to rates of methane oxidation. In my
experience, all FDR-correction is doing is removing variables (in this case,
taxa) that have p-values above some adjusted threshold that is determined by
the total number of variables in a model.

The 460 taxa were significant after FDR correction, not before. we have updated
the text to make this clearer. Indeed, FDR correction sets a threshold for
significance that is less conservative than Bonferroni correction, but
ultimately it is just a significance threshold. It is true that correcting for
multiple comparisons does not come from population genetics. We have removed
the idea of correcting for multiple comparisons from the three “take-home”
points and have focused our text on describing the other two points. We have
also tried to make it clear that the main take home lesson from
genotype-phenotype mapping is not any particular statistical approach, but
rather a general way of thinking about these statistical relationships.

-- The section on population stratification starting on line 152 is
interesting. This doesn’t seem to be something that microbial ecologists often
do. I wonder if it would be helpful to expand upon this some more and perhaps
even show more explicitly how such spurious correlations could contribute
inferences either using the case study (methane oxidation) or another cartoon
type example, perhaps as a separate Box.

We have provided some examples of spurious correlations (using the methane
oxidation example) and have expanded on the conceptual ideas behind
stratification.

-- The section on high affinity methane oxidation needs work. I have no problem
with the case study per se, but I think it needs to be reworked if it is mean
to serve as a general example of a pipeline for how one might use
genotype-phenotype mapping to explore BEF. The second and third paragraphs of
this section don’t particularly seem aligned with the objective of the
manuscript.

Much of this section has been reworked and shaped into a Results & Discussion
section to interpret each step in the process and to better illustrate how and
why to perform each part of the analysis.

-- In the end, as someone who has never done genotype-phenotype mapping (but as
someone who has thought about BEF relationships from a statistical perspective)
I didn’t really come away feeling like a I had a deeper understanding or
appreciation for how I might use the authors’ recommended approach, but I think
with some more effort and revision, this could be achieved. Since this seems
like a fairly novel idea and one that spans disciplines, I would encourage the
authors to think about how the paper might serve in part as a primer that would
help readers understand the conceptual and statistical analogies between
genotype-phenotype mapping and BEF research.

This has been addressed by many of the other comments, particularly in
expanding on the analogy and how some of the conceptual components might apply
to microbial communities. We have also tried to make the purpose of the example
clearer and interpret the results more thoroughly. Finally, we have added a
caveats/future directions section that provides some suggestions of what other
changes could be made to improve on this approach in future studies.

-- Thought based on and around text on lines 104. Examples of phenotypes like
height are classic example of a quantitative trait that is influenced by many
genes. Could statistical approaches from quantitative genetics be used in a BEF
context?

The way we think about this is that genotype-phenotype mapping is the purview
of population genetics. Quantitative genetics is simply a subset of population
genenetics that deals with continuous rather than discrete traits. Ecosystem
function is more in line with quantitative genetics, because we’re dealing with
quantitative traits of ecosystems (e.g, rates of fluxes of molecules), but
ultimately the statistical approaches are the same. We have used “population
genetics” throughout in order to consistently use one term and so the ideas
apply more generally.

Minor Comments

-- Line 40: Environmental DNA is introduced here without a description of the
type of sequencing data being discussed, which may be critical to understanding
the information gathered from the DNA, be it phylogenetic or functional-gene
markers. Later, in line 208, 16S sequences are discussed as a way to identify
organisms that does not provide functional information. Even later, in line
302, it is clarified that functional group abundance can be measured via
genetic markers from environmental DNA.

We revised line 40 to specifically describe the type of sequencing data (16S
sequences or metagenomic sequences or qPCR of functional genes). We revised
line 302 to specifically refer to qPCR of functional genes. In all cases, the
phrase “environmental DNA” has been replaced with the specific type of
sequencing data being discussed. Sections 1.2 and 2.1 expand on the idea behind
functional genes.

-- Lines 140-167: Consider splitting Section 1.6 into two sections, one on
multiple comparisons and one on covariation, to stay consistent with the
expected discussion of three insights as identified in the abstract.

We removed the multiple comparisons section as per above comments and created a
separate section regarding covariation.

-- Lines 150-151: Consider expanding upon commonly used corrections in GWAS
studies. An additional sentence of two about a couple common corrections may
make this section stand on its own and allow the methods to be clearer to the
reader.

We added a few sentences elaborating on PC correction and variance component
modeling. Also, we added text regarding the design of the experiment around a
homogeneous population, which is considered a good practice in GWAS.

-- Lines 203-204: An explanation of the importance of site ID and ecosystem
type in structuring communities would aid the reader in understanding what
Figure 2 adds to the example. Is Figure 2 included to show the reader that the
environmental variation, community structure and geographic distance structure
communities? In general, it wasn’t clear to me that Figures 1 and 2 were
critical or necessary to include.

We added more interpretation of Figure 2.

-- Lines 264-267: The proposed method for identifying taxa of small effect
could benefit from a clearer set-up. The idea is overshadowed by the discussion
of detecting loci of small effect in genotype-phenotype studies and needs a
stronger connection to microbial biodiversity ecosystem function relationships
to stand on its own.

We moved this up to section 2 and made it clearer how this would work in BEF
studies.

-- Lines 256-259: The idea of “uniquely associated” taxa should be introduced
earlier in the text, perhaps section 1.4. Without explaining the change in
question from associated to uniquely associated, the example loses some of its
power and the importance of genotype-phenotype mapping methods is undersold.

We added discussion of uniquely associated to the section on controlling for
population structure.

-- Lines 50-52: consider revising for clarity

This text has been rewritten and moved to the “Population Stratification”
section

-- Line 57. Typo at end of topic sentence.

The comma has been corrected to a period.

-- Line 262-263: Reword sentence for clarity

This text has been revised.

-- Line 235: Other statistical approaches used for understanding causal
relationships include structural equation modeling (SEM) and path analysis.
These techniques also provide opportunity to identify latent variables, i.e.,
variables that are likely important but are difficult to measure. Might such
approaches be mentioned, and could they be used in conjunction with findings
from genotype-phenotype mapping? For example, it would be impossible to build
an SEM with 406 taxa and make any sense of what was going on, but with a
reduced set of taxa (e.g., 6), a SEM becomes very tractable.

We have added some discussion of how to use the identified taxa including a
form of microbial polygenic risk scores and including the taxa in a structural
equation model (Section 3.3).

-- To what degree to microbial BEF studies suffer from the fact that we are all
using relative abundance data? This is an obstacle that genotype-phenotype
mapping doesn’t have to deal with, correct?

We added text describing this to the discussion.

-- No hyphen needed in “microbially-mediated”

We have changed all instances of “microbially-mediated” to “microbially
mediated.”

### 2019-12-10 manuscript/manuscript_rev1.Rmd
Update revision and reviewer response

library(tidyverse)
library(knitr)
library(flextable)
opts_chunk$set(echo = FALSE)

```

@@

Microbial physiological processes are intimately involved in nutrient cycling.
However, it remains unclear to what extent microbial diversity or community
composition is important for determining the rates of ecosystem-scale
functions. There are numerous examples where microbial diversity is correlated
with the rate of an ecosystem function, but exactly how microbial communities
“map” onto ecosystem functions remains unresolved.  This uncertainty limits our
ability to predict and manage crucial microbially mediated processes such as
nutrient loss and greenhouse gas emissions. Our lack of understanding stems
from the relatively large diversity of microorganisms, the difficulty in
directly identifying functional groups, and our limited ability to manipulate
microbial community attributes. For this reason, we propose that integrating
traditional biodiversity-ecosystem function research with ideas from
genotype-phenotype mapping could provide the new perspective our discipline
needs. We identify two insights from genotype-phenotype mapping that could be
useful for microbial biodiversity-ecosystem function studies: the concept of
“agnostic” mapping and controlling for population stratification to identify
microorganisms that uniquely contribute to ecosystem function. We illustrate
the potential for these approaches to elucidate microbial
biodiversity-ecosystem function relationships by analyzing a subset of
published data measuring methane oxidation rates from tropical soils. We assert
that combining the approaches of traditional biodiversity-ecosystem function
research with ideas from genotype-phenotype mapping will not only generate
novel hypotheses about how complex microbial communities drive ecosystem
function, but also help scientists predict and manage changes to ecosystem
functions resulting from human activities.

@@

Ecology is broadly focused on understanding biodiversity and how that
biodiversity shapes the ecosystems that humans depend on. Many ecosystem
processes essential to all of life are mediated by microorganisms and therefore
understanding the relationship between microbial biodiversity and ecosystem
function is important [@schimel1998, @singh2010]. Certain ecosystem functions
are correlated with microbial diversity, indicating that we should be able to
determine what aspects of microbial biodiversity influence ecosystem function.
However, attempts to describe that mapping have borne little fruit [@rocca2015,
@graham2016]. We argue that to overcome this challenge we should look to other
successful attempts at mapping biological variation onto higher order
processes. In particular, population genetics and the process of
genotype-phenotype mapping provide a number of potentially useful insights. For
example, genotype-phenotype mapping often makes few assumptions about the
nature of the map, i.e., it is “agnostic.” In addition, population geneticists
have developed rigorous methods for reducing potentially confounding
relationships such as geographic structuring of populations. Finally, the
ultimate goal of genotype-phenotype mapping is to identify the unique
contribution of genotype to phenotype separate from other drivers of phenotypic
variation such as the environment. Inspired by this, we suggest a reframe of
the question, "is microbial biodiversity related to ecosystem function?" to, "what
is the unique contribution of the microbiome to ecosystem function independent
of the environment?"

@@

There is increasing evidence for microbial biodiversity-ecosystem function
relationships. For example, there are positive correlations between microbial
diversity and ecosystem multifunctionality for a variety of ecosystems and most
major lineages of microorganisms [@jing2015, @delgado-baquerizo2016,
@delgado-baquerizo2017]. Adding microbial diversity or abundance to ecosystem
models can in some cases improve model accuracy [@graham2016]. Contrived
communities that vary in richness and communities created through sequential
dilution or varying filter size to generate differences in diversity can also
exhibit differences in ecosystem function [@philippot2013, @wagg2014, @schnyder2018,
@maron2018].  Finally, reciprocal transplant and common garden experiments that
manipulate the connection between community composition and environment reveal
differences in ecosystem function for communities of distinct
origins [@cavigelli2000, @strickland2009, @glassman2018]. Given these
relationships, many investigators have now moved on to the challenge of
identifying the aspects of microbial biodiversity (e.g., specific taxa, genes,
functional groups, etc.) that influence a given ecosystem function; however,
this has proven especially challenging.

@@

Most studies that attempt to identify the aspects of microbial biodiversity
that influence a given ecosystem function focus on “functional” gene or
transcript abundance.  In this case, qPCR or shotgun metagenomic sequencing is
used to estimate the abundance of a gene or transcript that is a putative
marker for a microbial process (and thus a marker for the functional group that
performs this process). For example, the gene _mcrA_, which encodes a subunit of
the enzyme that performs the final step in methanogenesis, is commonly used as
a marker for methanogenesis and for the methanogen functional group. Other
examples include _pmoA_ and methanotrophy, _nifH_ and nitrification, and _nosZ_ and
denitrification. The abundance of these markers is often hypothesized to be
predictive of the rate of their associated processes (for example, the
abundance of _mcrA_ is often hypothesized to be related to the rate of
methanogenesis).

Some ecosystem functions in certain ecosystems can be predicted from the
abundance or transcriptional activity of genetic markers for those functions.
For example, soil methane production and consumption can under some
circumstances be predicted from the genetic markers _mcrA_ and _pmoA_
[@freitag2009, @freitag2010, @mccalley2014]. However, for most ecosystem
functions the abundance of a functional gene or transcript is rarely positively
correlated with the rate of the corresponding process [@rocca2015]. The
cases where there is a positive correlation tend to be restricted to
agricultural ecosystems and for certain functions within the nitrogen cycle
[@rocca2015]. In general, including aspects of microbial biodiversity
(e.g., functional gene abundance or diversity) improve models of ecosystem
function less than one third of the time, and increase variance explained by an
average of only 8 percentage points over environmental variables [@graham2016].

@@

This kind of “many-to-one” mapping is analogous to the challenge of identifying
the genetic basis of complex traits in organismal populations. In such
“genotype-phenotype” mapping studies, a population exhibits variation in a
phenotype (e.g., height or disease state) as well as variation in potentially
thousands of single nucleotide polymorphisms (SNPs). To identify the genetic
basis for a trait, investigators sample from this population and correlate
phenotype with genotype. While some phenotypes (e.g., the propensity for
diseases such as Parkinson’s) are controlled by a single genetic locus
[@kerem1989, @macdonald1992], most traits depend on a large number of genes
that control variation in that phenotype [@hindorff2009, reich2001]. In
addition, there is often no a priori expectation about which regions of the
genome control that trait so we must search for genetic markers
throughout the genome. If a marker is significantly correlated with the
phenotype of interest, this either indicates it is inside of a gene with a
direct or indirect effect on phenotype or that it is in linkage disequilibrium
(i.e., non-random association between two alleles) with a causal gene.

There are a number of parallels between this challenge faced by organismal
biologists and that facing microbial community ecologists. They both involve
large numbers of statistical comparisons. Both are attempting to identify
causal relationships that are potentially confounded by complex patterns of
covariation. There is often no strong expectations about which entities (i.e.,
which genomic regions or which microbial genes or lineages) are most likely to
be causally-related to phenotype or function, and thus “agnostic” approaches
are needed. For some ecosystem functions, it is possible that a single
microbial taxon could substantially influence its rate. For example, methane
flux from permafrost in Sweden may be controlled by a single taxon [@mccalley2014]. But
for most ecosystem functions, there could be many taxa of small effect that
contribute to the rate of ecosystem function. Finally, both ultimately require
manipulation (of genes or taxa) to establish causation.

@@

Most microbial biodiversity-ecosystem function research up to this point has
used an approach analogous to that used by plant ecologists studying
biodiversity-ecosystem function relationships. This approach is to measure or
manipulate the diversity of a taxonomic group (e.g., plants) and look for an
association with the function performed by that group (e.g., primary
productivity). We can think of plants as a “functional group,” i.e., a group of
taxa united by their ability to perform a particular ecosystem function. For
microbes, estimating functional group abundance can be much more challenging.
From a small number of cultured isolates, we have a provisional understanding
of which microbes might be involved in some ecosystem functions. By sequencing
the genomes of these isolates, we have identified genetic markers for certain
functions, which we call “functional genes.” But most microbial taxa remain
uncultured and we do not know the function of most microbial taxa detected in
environmental samples [@hug2016, @martiny2019]. In addition, there have
been recent discoveries of functions in unexpected taxonomic groups, for
example methanogenesis by fungi and cyanobacteria, a function previously
considered restricted to the Archaea [@bizic-ionescu2019, @lenhart2012].

As illustrated earlier, these functional markers are not correlated with their
corresponding ecosystem function in most ecosystems and for most processes. In
addition, they provide little explanatory power to models of ecosystem
function. Because of this, it might be prudent to look more agnostically at
microbial communities to identify taxa, groups of taxa or genes that are
important for predicting the rates of ecosystem functions rather than assuming
that the genetic markers we have provisionally identified for a given function
represent the most likely taxa or genes involved. This agnostic approach is
analogous to the approach of many genotype-phenotype mapping studies (e.g.,
genome-wide association studies), which often look for associations between a
phenotype and loci anywhere in the genome.

Beyond finding new physiologies in unexpected places, there are other reasons
for looking agnostically. In the case of microbial functions, it may be that
the organism that performs a function is not the limiting factor for the rate
of that function. For example, the rate of soil-to-atmosphere methane flux
could be limited by methanogens or methanotrophs or the balance of the two. It
could also be limited by the bacteria that produce the fermentative byproducts
that methanogens depend on. Or there could be indirect limitation by organisms
that liberate nitrogen or phosphorus into mineral forms. In other words, the
influence of microbial communities on the rate ecosystem function could
represent a complex metabolic network much like the regulation of gene
expression in organisms that partially determine their phenotype. These broader
patterns of biodiversity-ecosystem function relationships would be invisible to
any study that focuses on putative functional groups without considering the
reads that underlie those functional groups.

@@

It is widely accepted that organisms, including microorganisms, exhibit
population stratification due to geographic and environmental separation
[@wright1943, @martiny2006]. This can lead to spurious associations between
phenotypes and genetic markers that are at high frequency in isolated
sub-populations. Association studies generally control for population
stratification by accounting for shared ancestry among organisms in a
population when modeling the connection between genotype and phenotype.
Typically, microbial biodiversity-ecosystem function studies do not account for
population stratification (i.e., community similarity among ecosystems),
although there are some exceptions [@meyer2019, @qin2012, @lloyd-price2019].
Community similarity (the community analogue of shared ancestry among
organisms) is not as tightly linked to geography or environment as is shared
ancestry, and it could be very useful to account for these separately in
microbial studies, especially if one is particularly interested in how
composition alters function independent of the underlying environmental
variation.

Genome-wide association studies correct for stratification using a variety of
methods. Generally, they ignore the underlying environmental and spatial
distance between samples and instead use shared ancestry as a proxy for local
selection and assortative mating. A common approach is to perform a regression
of phenotype and shared ancestry (computed as the first one or more principal
components of the genotype matrix) and then use the residuals from this model
as the values for phenotype in a subsequent regression using the genotypes
directly [@price2006]. This principal component correction is designed to test the
effect of individual genes after removing the effect of shared ancestry among
individuals. Another approach, employed in our example, is variance component
modeling (or mixed modeling, hierarchical modeling, etc.) where genotypic
similarity is included as a covariate in the model to control for
stratification while testing the genotype-phenotype connection [@kang2010].

If we control for covariates such as community similarity, geographic distance,
or environmental similarity, it changes the nature of our question. For
example, if we test the correlation between the relative abundance of a taxon
and the rate of methane flux we are asking “is this taxon correlated with
methane flux?” If we find a significant result, that may be because variation
in the abundance of that organism directly or indirectly contributes to methane
flux. However, it might also be that that organism lives only in ecosystems
that happen to have a high rate of ecosystem function. We would not be able to
distinguish these outcomes with this simple correlation. If we then include
environmental variables or environmental similarity as a covariate, we could
start to ask, “Is the taxon uniquely associated with function independent of
the environment?” By “uniquely associated”, we mean those taxa associated with
function irrespective of environmental conditions, local community structure,
or spatial proximity. This is not only a more specific question than is usually
asked in microbial biodiversity-ecosystem function studies, but it is also a
more appropriate one, especially if one is interested in how to incorporate
microbial community data into ecosystem models. Finally, it is interesting on
its own to understand whether microbial communities are selected by the
underlying environmental conditions to produce a particular rate of ecosystem
function (the conduit) or whether there is variation in community composition
independent of the environment that explains some variation in ecosystem
function.

@@

```{r taxa, out.width="768px", out.height="384px"}
knitr::include_graphics('../figures/effect_sizes.pdf')
```

>>>>

library(tidyverse)
library(knitr)
library(flextable)
opts_chunk$set(echo = FALSE) 
```

@@

Microbial physiological processes are intimately involved in nutrient cycling.
However, it remains unclear to what extent microbial diversity or community
composition is important for determining the rates of ecosystem-scale
functions. There are many examples of positive correlations between microbial
diversity and ecosystem function, but how microbial communities “map” onto
ecosystem functions remains unresolved. This uncertainty limits our
ability to predict and manage crucial microbially mediated processes such as
nutrient losses and greenhouse gas emissions. To overcome this challenge, we
propose integrating traditional biodiversity-ecosystem function research with
ideas from genotype-phenotype mapping in organisms. We identify two insights
from genotype-phenotype mapping that could be useful for microbial
biodiversity-ecosystem function studies: the concept of searching
“agnostically” for markers of ecosystem function and controlling for population
stratification to identify microorganisms uniquely associated with ecosystem
function. We illustrate the potential for these approaches to elucidate
microbial biodiversity-ecosystem function relationships by analyzing a subset
of published data measuring methane oxidation rates from tropical soils. We
assert that combining the approaches of traditional biodiversity-ecosystem
function research with ideas from genotype-phenotype mapping will generate
novel hypotheses about how complex microbial communities drive ecosystem
function and help scientists predict and manage changes to ecosystem functions
resulting from human activities.

# Keywords

Microbial ecology; Biodiversity-ecosystem function; Structure-function relationships;
Microbial processes; Microbial community composition; Biogeochemical cycles

@@

Ecology is broadly focused on understanding biodiversity and how that
biodiversity shapes the ecosystems that humans depend on. Many ecosystem
processes essential to all of life are mediated by microorganisms and therefore
understanding the relationship between microbial biodiversity and ecosystem
function is important [@schimel1998; @singh2010]. Certain ecosystem functions
are correlated with microbial diversity, indicating that we should be able to
determine what aspects of microbial biodiversity influence ecosystem function.
However, attempts to describe that mapping have borne little fruit [@rocca2015;
@graham2016]. We argue that to overcome this challenge we should look to other
successful attempts at mapping biological variation onto higher order
processes. In particular, population genetics and the process of
genotype-phenotype mapping provide a number of potentially useful insights. For
example, genotype-phenotype mapping often makes few assumptions about the
nature of the map, i.e., it is “agnostic.” In addition, population geneticists
have developed rigorous methods for reducing potentially confounding
relationships such as geographic structuring of populations. Finally, the
ultimate goal of genotype-phenotype mapping is to identify the unique
contribution of genotype to phenotype separate from other drivers of phenotypic
variation such as the environment. Inspired by this, we suggest a reframe of
the question, "is microbial biodiversity related to ecosystem function?" to, "what
is the unique contribution of the microbiome to ecosystem function independent
of the environment?"

@@

There is increasing evidence for microbial biodiversity-ecosystem function
relationships. For example, there are positive correlations between microbial
diversity and ecosystem multifunctionality for a variety of ecosystems and most
major lineages of microorganisms [@jing2015; @delgado-baquerizo2016;
@delgado-baquerizo2017]. Adding microbial diversity or abundance to ecosystem
models can in some cases improve model accuracy [@graham2016]. Contrived
communities that vary in richness and communities created through sequential
dilution or varying filter size to generate differences in diversity can also
exhibit differences in ecosystem function [@philippot2013; @wagg2014; @schnyder2018;
@maron2018].  Finally, reciprocal transplant and common garden experiments that
manipulate the connection between community composition and environment reveal
differences in ecosystem function for communities of distinct
origins [@cavigelli2000; @strickland2009; @glassman2018]. Given these
relationships, many investigators have now moved on to the challenge of
identifying the aspects of microbial biodiversity (e.g., specific taxa, genes,
functional groups, etc.) that influence a given ecosystem function; however,
this has proven especially challenging.

@@

Most studies that attempt to identify the aspects of microbial biodiversity
that influence a given ecosystem function focus on “functional” gene or
transcript abundance.  In this case, qPCR or shotgun metagenomic sequencing is
used to estimate the abundance of a gene or transcript that is a putative
marker for a microbial process (and thus a marker for the functional group that
performs that process). For example, the gene _mcrA_, which encodes a subunit of
the enzyme that performs the final step in methanogenesis, is commonly used as
a marker for methanogenesis and for the methanogen functional group. Other
examples include _pmoA_ and methanotrophy, _nifH_ and nitrification, and _nosZ_
and denitrification. It is often hypothesized that the abundance of these
markers is predictive of the rate of the associated processes (for example,
it is hypothesized that the abundance of _mcrA_ is related to the rate of
methanogenesis).

Some ecosystem functions in certain ecosystems can be predicted from the
abundance or transcriptional activity of genetic markers for those functions.
For example, soil methane production and consumption can under some
circumstances be predicted from the genetic markers _mcrA_ and _pmoA_
[@freitag2009; @freitag2010; @mccalley2014]. However, for most ecosystem
functions the abundance of a functional gene or transcript is rarely positively
correlated with the rate of the corresponding process [@rocca2015]. The
cases where there is a positive correlation tend to be restricted to
agricultural ecosystems and certain functions within the nitrogen cycle
[@rocca2015]. In general, including aspects of microbial biodiversity
(e.g., functional gene abundance or diversity) improves models of ecosystem
function less than one third of the time and increases variance explained by an
average of only 8 percentage points over environmental variables [@graham2016].

@@

This kind of “many-to-one” mapping is analogous to the challenge of identifying
the genetic basis of complex traits in organismal populations. In such
“genotype-phenotype” mapping studies, a population exhibits variation in a
phenotype (e.g., height or disease state) as well as variation in potentially
thousands of single nucleotide polymorphisms (SNPs). To identify the genetic
basis for a trait, investigators sample from this population and correlate
phenotype with genotype. While some phenotypes (e.g., the propensity for
diseases such as Parkinson’s) are controlled by a single genetic locus
[@kerem1989; @macdonald1992], most traits depend on a large number of genes
that control variation in phenotype [@hindorff2009; @reich2001]. In
addition, there is often no a priori expectation about which regions of the
genome control that trait so we must search for genetic markers
throughout the genome. If a marker is significantly correlated with the
phenotype of interest, this either indicates it is inside of a gene with a
direct or indirect effect on phenotype or that it is in linkage disequilibrium
(i.e., non-random association between two alleles) with a causal gene.

There are a number of parallels between this challenge faced by organismal
biologists and that facing microbial community ecologists. They both involve
large numbers of statistical comparisons. Both are attempting to identify
causal relationships that are potentially confounded by complex patterns of
covariation. There is often no strong expectation about which entities (i.e.,
which genomic regions or which microbial genes or lineages) are most likely to
be causally related to phenotype or function, and thus “agnostic” approaches
are needed. For some ecosystem functions, it is possible that a single
microbial taxon could substantially influence its rate. For example, methane
flux from permafrost in Sweden may be controlled by a single taxon [@mccalley2014]. But
for most ecosystem functions, there could be many taxa of small effect that
contribute to the rate of ecosystem function. Finally, both ultimately require
manipulation (of genes or taxa) to establish causation.

@@

Most microbial biodiversity-ecosystem function research up to this point has
used an approach analogous to that used by plant ecologists studying
biodiversity-ecosystem function relationships. This approach is to measure or
manipulate the diversity of a taxonomic group (e.g., plants) and look for an
association with the function performed by that group (e.g., primary
productivity). We can think of plants as a “functional group,” i.e., a group of
taxa united by their ability to perform a particular ecosystem function. For
microbes, estimating functional group abundance can be much more challenging.
From a small number of cultured isolates, we have a provisional understanding
of which microbes might be involved in some ecosystem functions. By sequencing
the genomes of these isolates, we have identified genetic markers for certain
functions, which we call “functional genes.” But most microbial taxa remain
uncultured and we do not know the function of most microbial taxa detected in
environmental samples [@hug2016; @martiny2019]. In addition, there have
been recent discoveries of functions in unexpected taxonomic groups, for
example methanogenesis by fungi and cyanobacteria, a function previously
considered restricted to the archaea [@bizic-ionescu2019; @lenhart2012].

As illustrated earlier, these functional markers are not correlated with their
corresponding ecosystem function in most ecosystems and for most processes. In
addition, they provide little explanatory power to models of ecosystem
function. Because of this, it might be prudent to look more agnostically at
microbial communities to identify taxa, groups of taxa, or genes that are
important for predicting the rates of ecosystem functions rather than assuming
that the genetic markers we have provisionally identified for a given function
represent the most likely taxa or genes involved. This agnostic approach is
analogous to the approach of many genotype-phenotype mapping studies (e.g.,
genome-wide association studies), which often look for associations between a
phenotype and loci anywhere in the genome.

Beyond finding new physiologies in unexpected lineages, there are other reasons
for looking agnostically. In the case of microbial functions, it may be that
the organism that performs a function is not the limiting factor for the rate
of that function. For example, the rate of soil-to-atmosphere methane flux
could be limited by methanogens or methanotrophs or the balance of the two. It
could also be limited by the bacteria that produce the fermentative byproducts
that methanogens use as substrates. Or there could be indirect limitation by organisms
that liberate nitrogen or phosphorus into mineral forms. In other words, the
influence of microbial communities on the rate of ecosystem function could
represent a complex metabolic network much like the regulation of gene
expression in organisms that partially determines their phenotype. These broader
patterns of biodiversity-ecosystem function relationships would be invisible to
any study that solely focuses on the most relevant functional group without
considering the possible influence of other taxa.

@@

It is widely accepted that organisms, including microorganisms, exhibit
population stratification due to geographic and environmental separation
[@wright1943; @martiny2006]. This can lead to spurious associations between
phenotypes and genetic markers that are at high frequency in isolated
sub-populations. Association studies generally control for population
stratification by accounting for shared ancestry among organisms in a
population when modeling the connection between genotype and phenotype.
Typically, microbial biodiversity-ecosystem function studies do not account for
population stratification (i.e., community similarity among ecosystems),
although there are some exceptions [@meyer2019; @qin2012; @lloyd-price2019].
Community similarity (the community analogue of shared ancestry among
organisms) is not as tightly linked to geography or environment as is shared
ancestry. Therefore, it could be useful to account for these separately in
microbial studies, particularly if one is interested in quantifying the effect
of microbial communities on ecosystem function independent of these other factors.

Genome-wide association studies correct for stratification using a variety of
methods. Generally, they ignore the underlying environmental and spatial
distance between samples and instead use shared ancestry as a proxy for local
selection and assortative mating. A common approach is to perform a regression
of phenotype and shared ancestry (computed as the first one or more principal
components of a genotype matrix) and then use the residuals from this model
as the values for phenotype in a subsequent regression using the genotypes
directly [@price2006]. This principal component correction is designed to test the
effect of individual genes after removing the effect of shared ancestry among
individuals. Another approach, employed in our example, is variance component
modeling (or mixed modeling, hierarchical modeling, etc.) where genotypic
similarity is included as a covariate in the model to control for
stratification while testing the genotype-phenotype connection [@kang2010].

If we control for covariates such as community similarity, geographic distance,
or environmental similarity, it changes the nature of our question. For
example, if we test the correlation between the relative abundance of a taxon
and the rate of methane flux we are asking “is this taxon correlated with
methane flux?” If we find a significant result, that may be because variation
in the abundance of that organism directly or indirectly contributes to methane
flux. However, it might also be that that organism lives only in ecosystems
that happen to have a high rate of methane flux. In this scenario, we are
unable to distinguish these possibilities. However, if we add environmental
variables or environmental similarity as a covariate in our model, we can ask,
“Is this taxon uniquely associated with function in a way that it is
independent of the environment?” By “uniquely associated,” we mean those taxa
associated with function irrespective of environmental conditions, local
community structure, or spatial proximity. This slight reframing of the
question could be especially rewarding for microbial biodiversity-ecosystem
function research, particularly as it relates to incorporating microbial
community data into ecosystem models.  Finally, it is interesting in its own right to
understand whether microbial communities are selected by the underlying
environmental conditions to produce a particular rate of ecosystem function or
whether community variation has functional consequences independent of the
environment.

@@

```{r taxa, out.width="768px", out.height="384px", include=FALSE}
knitr::include_graphics('../figures/effect_sizes.pdf')
```
